---
title: "b. Added value"
output:
  rmarkdown::html_vignette:
#    highlight: "declared.theme"
#    css: "declared.css"
vignette: >
  %\VignetteIndexEntry{b. Added value}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r vignettesetup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(haven)
set.seed(12345)
```


```{r}
library(declared)

x2 <- declared(
  x = c(1:5, -91),
  labels = c("Missing" = -91),
  na_value = -91
)
```

The proposed method to declare missing values is unique in the R ecosystem.
Differentiating between empty and declared missing values opens the door to a
new set of challenges for which base the R does not have built-in functionality.

For instance, the declared missing values can be compared against both the
original values and their labels:

```{r}
x2 == -91
x2 == "Missing"
```

Similar methods have been added to the primitive functions `"<"`{.R}, `">"`{.R} and
`"!="`{.R} etc., to allow a fully functional collection of subsetting possibilities.

Combining on this type of vector creates an object of the same class:

```{r}
x2 <- c(x2, -91)
x2
```

Most functions are designed to be as similar as possible, for instance,
`labels()`{.R} to retrieve or add/change value labels:

```{r}
labels(x2) <- c("Does not know" = -92, "Not responded" = -91)
x2
```

The value `-92` is now properly labelled, and can further be declared missing. Such declarations do not necessarily have to use the main function
`declared()`{.R}, due to the separate functions `missing_values()`{.R} and
`missing_range()`{.R}:

```{r}
missing_values(x2) <- c(-91, -92)
missing_range(x2) <- c(-91, -95)
x2
```

To ease the smooth inter-operation with packages **haven** and **labelled**, the
following functions are of interest: `undeclare()`{.R}, `as.haven()`{.R} and
`as.declared()`{.R}.

The function `undeclare()`{.R} replaces the NAs with their declared missing values.
The result is still an object of class `"declared"`{.R}, but all missing values (and
missing range) are stripped off the vector, and values are presented as they have
been collected. All other attributes of interest (variable and value labels) are
retained and printed accordingly. Activating the argument `drop`{.R} eliminates all
classes and attributes, returning a regular R object:

```{r undeclare}
undeclare(x2, drop = TRUE)
```

The function `as.haven()`{.R} coerces the resulting object to the class
`"haven_labelled_spss"`{.R}, and the function `as.declared()`{.R} reverses the process:

```{r haven}
xh <- as.haven(x2)
xh

as.declared(xh)
```

The missing values are properly formatted, even inside the base data frame:

```{r dataframe}
dfm <- data.frame(x1 = letters[1:7], x2)
dfm
```

If users prefer a tibble instead of a data frame, the objects of class
"`declared"`{.R} are properly formatted in a similar way to those from package
**haven**:

```{r tibble}
tibble::as_tibble(dfm)
```

Special challenges are associated with sorting and ordering the declared
objects, where not all missing values are treated the same.

```{r sort}
x3 <- declared(
  x = c(1:5, -91, NA, -92, -91),
  na_value = c(-92, -91)
)
sort(x3, na.last = TRUE)
```

Sorting in decreasing order applies the same order to the missing values:

```{r}
sort(x3, na.last = TRUE, decreasing = TRUE)
```

This custom function benefits from an additional argument `empty.last`{.R}
(internally passed to the ordering function), to allow sorting within the
missing values:

```{r}
sort(x3, na.last = TRUE, decreasing = TRUE, empty.last = FALSE)
```

All types of variables (categorical and numerical) can have declared missing
values. There are situations when values are not missing randomly but with a specific reason. In social research, respondents often can not or do not want to provide an answer for a certain question, be it categories of opinions or pure numerical answers like age, income, etc.

Base R has a clear distinction between numerical and categorical variables. For the latter, R provides a special type of object called `factor`{.R}.
The following object simulates such a categorical variable, for instance
political orientation:

```{r}
x4 <- declared(
  x = c(1:3, -91),
  labels = c("Left" = 1, "Middle" = 2, "Right" = 3, "Apolitic" = -91),
  na_value = -91,
  label = "Respondent's political orientation"
)

x4
```

Such a variable could in principle be constructed directly as a factor:

```{r}
x5 <- factor(
  c("Left", "Middle", "Right", "Apolitic"),
  levels = c("Left", "Middle", "Right", "Apolitic")
)

x5
```

The base factors provide no possibility to assign specific values for the specific categories, and most importantly, they do not differentiate between valid values and (declared) missing values. To make the functionality consistent with the base
treatment of NA values, coercing to factors defaults to dropping the declared
missing values because of the default value of the argument `drop_na`:

```{r}
as.factor(x4)
as.factor(x4, drop_na = TRUE)

# essentially acting as:
as.factor(drop_na(x4))
```

Converting the declared missing values to factor levels is something for which the
function `undeclare()`{.R} proves useful, and switching between factors and 
declared objects, preserving the declared missing values, is now straightforward:

```{r}
as.factor(undeclare(x4))
```

This is almost identical, but differs from `x5` with respect to the level orders.
It happens because the missing value -91, coerced to a valid value, becomes the
first in the order of the categories while normally, a categorical variable displays
the valid values first, and the missing values last.

If the original order is important, the argument `drop_na` can be deactivated:

```{r}
as.factor(x4, drop_na = FALSE)
```



The reverse process is also possible, to convert / coerce factors to declared
labelled objects:


```{r}
as.declared(x5, na_values = 4)
```


When the declared objects are constructed as categorical variables to replace
the base factors, the function `as.character()`{.R} extracts the categories
in a similar way to factors, via a dedicated class method for declared objects:

```{r}
as.character(x4)
```

Similarly, the function `as.character()`{.R} drops the declared missing values
by default, something that can be prevented by either of the following two
commands:

```{r}
as.character(undeclare(x4))

as.character(x4, drop_na = FALSE)
```

The base factors store the categories using numeric values, which is also the most often scenario for the declared objects. But this is not at all mandatory, as the declared objects are also able to ingest character objects:

```{r}
x6 <- declared(
  x = sample(
    c("a", "b", "c", "z"),
    20,
    replace = TRUE
  ),
  labels = c("Left" = "a", "Middle" = "b", "Right" = "c", "Apolitic" = "z"),
  na_values = "z"
)

x6
```

Either as a character, numeric or categorical, it is possible to declare and use
special types of missing values, employing this new object type of class
`"declared"`{.R}.

Factors and `haven`{.R} objects have default coercion methods, but not all types
of objects can be automatically coerced to this class. To meet this possibility,
the main functions `declared()`{.R}, `as.declared()`{.R}, `undeclare()`{.R} and
`drop_na()`{.R} are all generic, allowing full flexibility for any other package
to create custom (coercion) methods for different classes of objects, thus
facilitating and encouraging a widespread use.
