---
title: "c. Weighting (declared) values"
output:
  rmarkdown::html_vignette:
#    highlight: "declared.theme"
#    css: "declared.css"
vignette: >
  %\VignetteIndexEntry{c. Weighting (declared) values}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


For the examples in this vignette, the following data frame is created:

```{r}
library(declared)

n <- 1234
set.seed(n)
dfm <- data.frame(
  Area = declared(
    sample(1:2, n, replace = TRUE, prob = c(0.45, 0.55)),
    labels = c("Rural" = 1, "Urban" = 2)
  ),
  Gender = declared(
    sample(1:2, n, replace = TRUE, prob = c(0.55, 0.45)),
    labels = c("Males" = 1, "Females" = 2)
  ),
  Opinion = declared(
    sample(c(1:5, NA, -91), n, replace = TRUE),
    labels = c(
      "Very bad" = 1, "Bad" = 2, "Neither" = 3,
      "Good" = 4, "Very good" = 5, "Don't know" = -91
    ),
    na_values = -91
  ),
  Age = sample(18:90, n, replace = TRUE),
  Children = sample(0:5, n, replace = TRUE)
)
```

One of the most interesting applications to make use of the declared missing
values are the tables of frequencies. The base function `table()`{.R} ignores
missing values by default, but they can be revealed by using the `useNA`{.R}
argument:

```{r}
table(dfm$Opinion, useNA = "ifany")
```

However, it does not differentiate between empty and declared missing values.
Since "Opinion" is the equivalent of a categorical variable, this can be
improved through a custom built coercion to the base `factor`{.R} class:

```{r}
table(as.factor(undeclare(dfm$Opinion)), useNA = "ifany")
```

The dedicated function `w_table()`{.R} does the same thing by automatically
recognizing objects of class `"declared"`{.R}, additionally printing more detailed
information:

```{r}
w_table(dfm$Opinion, values = TRUE)
```

The prefix `w_`{.R} from the function name stands for "weighted", this being another
example of functionality where the declared missing values play a different role
than the empty, base NA missing values.

It is important to differentiate between frequency weights, on one hand, and
other probability based, post-stratification weights on one other, the later
being thoroughly treated by the specialized package **survey**. The `w_`{.R} family
of functions are solely dealing with frequency weights, to allow corrections in
descriptive statistics, such as the tables of frequencies and other similar
descriptive measures for both categorical and numeric variables.

To exemplify, a frequency weights variable is constructed, to correct for the
distributions of gender by males and females, as well as the theoretical
distribution by residential areas differentiating between urban and rural
settlements.

```{r}
# Observed proportions
op <- proportions(with(dfm, table(Gender, Area)))

# Theoretical / population proportions:
# 53% Rural, and 50% Females
weights <- rep(c(0.53, 0.47), each = 2) * rep(0.5, 4) / op

dfm$fweight <- weights[
  match(10 * dfm$Area + dfm$Gender, c(11, 12, 21, 22))
]
```

The updated frequency table, this time using the frequency weights, can be
constructed by passing the weights to the argument `wt`{.R}:

```{r}
with(dfm, w_table(Opinion, wt = fweight, values = TRUE))
```

Except for the empty NA values, for which the weights cannot be applied, almost
all other frequencies (including the one for the declared missing value -91) are
now updated by applying the weights. This shows that, despite being interpreted
as "missing" values, the declared ones can and should also be weighted, with a
very useful result. Other versions of weighted frequencies do exist in R, but a
custom one was needed to identify (and weight) the declared missing values.

In the same spirit, many other similar functions are provided such as
`w_mean()`{.R}, `w_var()`{.R}, `w_sd()`{.R} etc., and the list will likely grow
in the future. They are similar to the base package counterparts, with a single
difference: the argument `na.rm`{.R} is activated by default, with or without
weighting. This is an informed decision about which users are alerted in the
functions' respective help pages.

The package **declared** was built with the specific intention to provide a
lightweight, zero dependency resource in the R ecosystem. It contains an already
extensive, robust and ready to use functionality that duly takes into account
the difference between empty and declared missing values.

It extends base R and opens up data analysis possibilities without precedent. By
providing generic classes for all its objects and functions, package
**declared** is easily extensible to any type of object, for both creation and
coercion to class `"declared"`{.R}.
